---
title: "Final Project"
format: pdf
editor: source
author: "Group 4"
---

## MEIN40330 - Final Project -- Question 4:

#### Can you suggest potential new drug targets using LINCS data? Or based on DepMap data? Or both?

Objective - identify genes or pathways that are potential drug targets for LUAD, meaning they are critical for cancer cell survival / proliferation (from DepMap), show altered expression / mutation in LUAD patients (from TCGA-LUAD), and/or exhibit functional relevance when perturbed (from LINCS)

#### Drug Target Criteria

-   genes with high expression or frequent mutations in TCGA-LUAD (indicating relevance to LUAD biology).
-   genes with high dependency scores in LUAD cell lines in DepMap (indicating essential for cancer cell survival)
-   genes whose perturbation in LINCS data alters LUAD-relevant phenotypes (e.g., reduced cell viability or proliferation).


# Setup for Data Analysis

```{r setup, message=FALSE, warning=FALSE}
#| message: false
#| warning: false

# Install and load pacman for efficient package management
if (!requireNamespace("pacman", quietly = TRUE)) {
  install.packages("pacman")
}
library(pacman)

# Install BiocManager if not already installed
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

# Define required CRAN and Bioconductor packages
cran_packages <- c("dplyr", "tibble", "ggplot2", "readr", "tidyr", "glmnet", "survival", "survminer", "data.table")
bioc_packages <- c("TCGAbiolinks", "survcomp", "cmapR", "rhdf5", "DESeq2", "org.Hs.eg.db", "SummarizedExperiment", "AnnotationDbi")

# Install CRAN packages if not already installed
pacman::p_load(char = cran_packages)

# Install Bioconductor packages if not already installed
for (pkg in bioc_packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    BiocManager::install(pkg, update = FALSE, ask = FALSE)
  }
}

# Install cmapR from GitHub if not already installed
if (!requireNamespace("cmapR", quietly = TRUE)) {
  pacman::p_load(devtools)
  devtools::install_github("cmap/cmapR")
}

# Load all required packages
pacman::p_load(char = cran_packages)
for (pkg in bioc_packages) {
  library(pkg, character.only = TRUE)
}
```

# Data Download and Preprocessing

The following chunk downloads raw TCGA-LUAD data.

```{r download_rawdata, message=FALSE, warning=FALSE, cache=TRUE}
#| message: false
#| warning: false
#| cache: true

# Section 1: Download TCGA-LUAD gene expression data (run only if not already downloaded)
setwd("/Users/martha/projects/lung-cancer-ODE/notebooks/q4")
data_dir <- "../../dat"
data_file <- file.path(data_dir, "luad_expr.rds")
tcga_degs_file <- file.path(data_dir, "tcga_genes_df.rds")
tcga_res_file <- file.path(data_dir, "tcga_res.rds")

if (!file.exists(tcga_degs_file) || !file.exists(tcga_res_file)) {
  if (!file.exists(data_file)) {

    # Query TCGA-LUAD dataset
    query_expr <- TCGAbiolinks::GDCquery(
      project = "TCGA-LUAD",
      data.category = "Transcriptome Profiling",
      data.type = "Gene Expression Quantification",
      workflow.type = "STAR - Counts"
    )

    # Download data
    TCGAbiolinks::GDCdownload(query_expr)
    
    # Prepare data
    luad_expr <- TCGAbiolinks::GDCprepare(query_expr)

    # Save to avoid redundant downloads
    saveRDS(luad_expr, file = data_file)
  } else {
    # Load cached data
    luad_expr <- readRDS(data_file)
  }

  # Section 2: Filter for tumor vs. normal samples
  sample_types <- c("Primary Tumor", "Solid Tissue Normal")
  counts <- SummarizedExperiment::assay(luad_expr)
  colData <- SummarizedExperiment::colData(luad_expr)
  keep <- colData$sample_type %in% sample_types

  # Verify that samples are available
  if (sum(keep) == 0) {
    stop("No samples found for specified sample types: ", paste(sample_types, collapse = ", "))
  }

  counts <- counts[, keep]
  colData <- colData[keep, ]

  # Section 3: Differential Expression Analysis
  dds <- DESeq2::DESeqDataSetFromMatrix(
    countData = round(counts),
    colData = colData,
    design = ~ sample_type
  )
  dds <- DESeq2::DESeq(dds)
  res <- DESeq2::results(dds, contrast = c("sample_type", "Primary Tumor", "Solid Tissue Normal"))

  # Section 4: Filter significant DEGs
  tcga_degs <- rownames(res)[!is.na(res$padj) & res$padj < 0.05 & abs(res$log2FoldChange) > 1]

  # Section 5: Map Ensembl IDs to gene symbols
  tcga_degs_clean <- sub("\\..*", "", tcga_degs) # Strip version numbers
  tcga_genes <- AnnotationDbi::mapIds(
    org.Hs.eg.db,
    keys = tcga_degs_clean,
    keytype = "ENSEMBL",
    column = "SYMBOL",
    multiVals = "first"
  )

  # Section 6: Summarize results
  # Convert to data frame for cleaner output
  tcga_genes_df <- data.frame(
    Ensembl = tcga_degs_clean,
    Symbol = tcga_genes,
    stringsAsFactors = FALSE
  )
  
  saveRDS(tcga_genes_df, file = tcga_degs_file)
  saveRDS(res, file = tcga_res_file)
} else {
  tcga_genes_df <- readRDS(tcga_degs_file)
  res <- readRDS(tcga_res_file)
}

# Display top 10 genes (or fewer if less available)
knitr::kable(head(tcga_genes_df, 10), caption = "Top 10 Differentially Expressed Genes (TCGA-LUAD)")

# Report total number of DEGs
cat("Total number of differentially expressed genes:", length(tcga_genes_df$Symbol[!is.na(tcga_genes_df$Symbol)]), "\n")
``` 

## Metadata Loading and Filtering

This section loads metadata from the LINCS dataset (GSE92742), validates the data, and filters for lung cancer cell lines and shRNA/drug perturbation signatures.

```{r, load_metadata, message=FALSE, warning=FALSE, cache=TRUE}
#| message: false
#| warning: false
#| cache: true

# Section 1: Define file paths

# Unzip files
system("gunzip ../../dat/*.gz")

files <- c(
  gene_info = file.path(data_dir, "GSE92742_Broad_LINCS_gene_info.txt"),
  cell_info = file.path(data_dir, "GSE92742_Broad_LINCS_cell_info.txt"),
  pert_info = file.path(data_dir, "GSE92742_Broad_LINCS_pert_info.txt"),
  sig_info = file.path(data_dir, "GSE92742_Broad_LINCS_sig_info.txt")
)

# Section 2: Validate files exist
for (file in files) {
  if (!file.exists(file)) {
    stop("File not found: ", file)
  }
}

# Section 3: Clear memory to free up space
rm(list = setdiff(ls(), c("data_dir", "files", "tcga_degs_file", "tcga_res_file")))
gc()

# Section 4: Load metadata
gene_info <- data.table::fread(files["gene_info"], sep = "\t", verbose = FALSE, showProgress = FALSE)
cell_info <- data.table::fread(files["cell_info"], sep = "\t", verbose = FALSE, showProgress = FALSE)
pert_info <- data.table::fread(files["pert_info"], sep = "\t", verbose = FALSE, showProgress = FALSE)
sig_info <- data.table::fread(
  files["sig_info"],
  sep = "\t",
  select = c("sig_id", "cell_id", "pert_id", "pert_type", "pert_iname"),
  verbose = FALSE,
  showProgress = FALSE
)

# Convert data.tables to data.frames for dplyr compatibility
gene_info <- as.data.frame(gene_info)
cell_info <- as.data.frame(cell_info)
pert_info <- as.data.frame(pert_info)
sig_info <- as.data.frame(sig_info)

# Section 5: Validate required columns are present
required_cols <- list(
  cell_info = "primary_site",
  sig_info = c("sig_id", "cell_id", "pert_id", "pert_type", "pert_iname")
)
for (df_name in names(required_cols)) {
  df <- get(df_name)
  missing_cols <- setdiff(required_cols[[df_name]], colnames(df))
  if (length(missing_cols) > 0) {
    stop("Missing required columns in ", df_name, ": ", paste(missing_cols, collapse = ", "))
  }
}

# Section 6: Filter lung cancer cell lines -- Examples: A549, NCIH1299, HCC827
lung_cell_lines <- cell_info %>%
  dplyr::filter(grepl("lung", tolower(primary_site))) %>%
  dplyr::pull(cell_id)

# Verify that lung cell lines are found 
if (length(lung_cell_lines) == 0) {
  stop("No lung cancer cell lines found in cell_info")
}

# Section 7: Filter signatures for lung cancer cell lines and shRNA/drug perturbations
lung_signatures <- sig_info %>%
  dplyr::filter(cell_id %in% lung_cell_lines & pert_type %in% c("trt_sh", "trt_cp")) %>%
  dplyr::select(sig_id, cell_id, pert_id, pert_type, pert_iname)

# Verify that signatures are found
if (nrow(lung_signatures) == 0) {
  stop("No signatures found for lung cancer cell lines with shRNA or drug perturbations")
}

# Section 8: Save filtered signatures to avoid reloading
saveRDS(lung_signatures, file = file.path(data_dir, "lung_signatures.rds"))

# Section 9: Display results
knitr::kable(head(lung_cell_lines, 10), col.names = "Lung Cancer Cell Lines", caption = "Top 10 Lung Cancer Cell Lines")
cat("Total number of lung cancer cell lines:", length(lung_cell_lines), "\n")
knitr::kable(head(lung_signatures, 10), caption = "Top 10 Filtered Signatures (Lung Cancer, shRNA/Drug)")
cat("Total number of filtered signatures:", nrow(lung_signatures), "\n")
```

# LINCS Level 5 Data Processing

This section loads LINCS Level 5 z-score data for lung cancer signatures, maps Entrez IDs to gene symbols, filters for shRNA and drug perturbations (gefitinib, erlotinib), and identifies significant genes with |z-score| > 2.

```{r, load_level_five_data, message=FALSE, warning=FALSE, cache=TRUE}
#| message: false
#| warning: false
#| cache: true

# Section 1: Define data directory and file paths
data_dir <- "../../dat"
gctx_file <- file.path(data_dir, "GSE92742_Broad_LINCS_Level5_COMPZ.MODZ_n473647x12328.gctx")
lung_sig_file <- file.path(data_dir, "lung_signatures.rds")

# Section 2: Validate file existence
if (!file.exists(gctx_file)) {
  stop("GCTX file not found: ", gctx_file)
}
if (!file.exists(lung_sig_file)) {
  stop("Lung signatures file not found: ", lung_sig_file)
}

# Section 3: Load lung signatures
lung_signatures <- readRDS(lung_sig_file)
lung_sig_ids <- lung_signatures$sig_id

# Section 4: Clear memory to free up space
rm(list = setdiff(ls(), c("data_dir", "gctx_file", "lung_signatures", "lung_sig_ids", "gene_info", "tcga_genes_df")))
gc()

# Section 5: Load Level 5 data for lung cancer signatures
ds_file <- file.path(data_dir, "lung_level5_data.rds")
if (!file.exists(ds_file)) {
  ds <- cmapR::parse_gctx(gctx_file, cid = lung_sig_ids)
  saveRDS(ds, file = ds_file)
} else {
  ds <- readRDS(ds_file)
}

# Section 6: Validate signature IDs
if (!all(lung_sig_ids %in% ds@cid)) {
  stop("Some lung signature IDs not found in GCTX file")
}

# Section 7: Map Entrez IDs to gene symbols
gene_ids <- ds@rid
gene_symbols <- gene_info %>%
  dplyr::filter(pr_gene_id %in% gene_ids) %>%
  dplyr::select(pr_gene_id, pr_gene_symbol)
if (nrow(gene_symbols) == 0) {
  stop("No matching gene IDs found in gene_info")
}

# Section 8: Filter shRNA and drug signatures
shRNA_signatures <- lung_signatures %>%
  dplyr::filter(pert_type == "trt_sh")
shRNA_sig_ids <- shRNA_signatures$sig_id
drug_signatures <- lung_signatures %>%
  dplyr::filter(pert_type == "trt_cp" & pert_iname %in% c("gefitinib", "erlotinib"))
drug_sig_ids <- drug_signatures$sig_id

# Verify signatures
if (length(shRNA_sig_ids) == 0) {
  stop("No shRNA signatures found")
}
if (length(drug_sig_ids) == 0) {
  stop("No drug signatures found for gefitinib or erlotinib")
}

# Section 9: Subset GCT for shRNA and drug signatures
shRNA_ds <- cmapR::subset_gct(ds, cid = shRNA_sig_ids)
drug_ds <- cmapR::subset_gct(ds, cid = drug_sig_ids)

# Save subsets to avoid reprocessing
saveRDS(shRNA_ds, file = file.path(data_dir, "shRNA_lung_signatures.rds"))
saveRDS(drug_ds, file = file.path(data_dir, "drug_lung_signatures.rds"))

# Section 10: Identify significant genes from shRNA signatures
z_scores <- shRNA_ds@mat
significant_genes <- apply(z_scores, 1, function(x) any(abs(x) > 2))
sig_gene_ids <- shRNA_ds@rid[significant_genes]
sig_gene_symbols <- gene_info %>%
  dplyr::filter(pr_gene_id %in% sig_gene_ids) %>%
  dplyr::pull(pr_gene_symbol)
if (length(sig_gene_symbols) == 0) {
  stop("No significant genes found with |z-score| > 2")
}

# Section 11: Display results
knitr::kable(head(gene_symbols, 10), caption = "Top 10 Gene ID to Symbol Mappings")
cat("Total number of mapped genes:", nrow(gene_symbols), "\n")
knitr::kable(head(shRNA_signatures, 10), caption = "Top 10 shRNA Signatures")
cat("Total number of shRNA signatures:", nrow(shRNA_signatures), "\n")
knitr::kable(head(drug_signatures, 10), caption = "Top 10 Drug Signatures (Gefitinib/Erlotinib)")
cat("Total number of drug signatures:", nrow(drug_signatures), "\n")
knitr::kable(head(data.frame(Gene_Symbol = sig_gene_symbols), 10), caption = "Top 10 Significant Genes (|z-score| > 2)")
cat("Total number of significant genes:", length(sig_gene_symbols), "\n")
# Candidate genes are: "DDR1"   "PAX8"   "GUCA1A" "EPHB3"  "ESRRA"  "TRADD"
```

# DepMap Data Processing

This section loads DepMap CRISPR gene effect and model data, filters for lung adenocarcinoma (LUAD) cell lines, identifies essential genes (CERES < -0.5 in 60–90% of LUAD cell lines), and intersects with TCGA and LINCS results to identify candidate target genes.

```{r, filter_depmap, message=FALSE, warning=FALSE, cache=TRUE}
#| message: false
#| warning: false
#| cache: true

# Section 1: Define file paths
files <- c(
  gene_effect = file.path(data_dir, "CRISPRGeneEffect.csv"),
  model_data = file.path(data_dir, "Model.csv")
)

# Section 2: Validate file existence
for (file in files) {
  if (!file.exists(file)) {
    stop("File not found: ", file)
  }
}

# Section 3: Clear memory to free up space
rm(list = setdiff(ls(), c("data_dir", "files", "tcga_degs_file", "tcga_res_file", "sig_gene_symbols")))
gc()

# Section 4: Load DepMap data
gene_effect <- data.table::fread(files["gene_effect"], verbose = FALSE, showProgress = FALSE)
model_data <- data.table::fread(files["model_data"], verbose = FALSE, showProgress = FALSE)
gene_effect <- as.data.frame(gene_effect)
model_data <- as.data.frame(model_data)

# Section 5: Rename first column of gene_effect to DepMap_ID
colnames(gene_effect)[1] <- "DepMap_ID"

# Section 6: Validate required columns
required_cols <- list(
  gene_effect = "DepMap_ID",
  model_data = c("ModelID", "CellLineName", "OncotreeLineage", "OncotreeSubtype")
)
for (df_name in names(required_cols)) {
  df <- get(df_name)
  missing_cols <- setdiff(required_cols[[df_name]], colnames(df))
  if (length(missing_cols) > 0) {
    stop("Missing required columns in ", df_name, ": ", paste(missing_cols, collapse = ", "))
  }
}

# Section 7: Filter for LUAD cell lines
luad_cells <- model_data %>%
  dplyr::filter(grepl("lung", tolower(OncotreeLineage)) &
                grepl("adenocarcinoma|non-small cell lung", tolower(OncotreeSubtype))) %>%
  dplyr::pull(ModelID)
if (length(luad_cells) == 0) {
  stop("No LUAD cell lines found in model_data")
}

# Section 8: Subset gene effect data for LUAD cell lines
luad_gene_effect <- gene_effect %>%
  dplyr::filter(DepMap_ID %in% luad_cells)
if (nrow(luad_gene_effect) == 0) {
  stop("No gene effect data found for LUAD cell lines")
}

# Section 9: Identify essential genes (CERES < -0.5 in 60–90% of LUAD cell lines)
essential_genes <- luad_gene_effect %>%
  dplyr::select(-DepMap_ID) %>%
  dplyr::summarise_all(~ mean(. < -0.5, na.rm = TRUE)) %>%
  tidyr::pivot_longer(everything(), names_to = "gene", values_to = "prop_essential") %>%
  dplyr::filter(prop_essential > 0.6 & prop_essential < 0.9) %>%
  dplyr::mutate(gene_symbol = sub(" \\(.*\\)", "", gene)) %>%
  dplyr::distinct(gene_symbol, .keep_all = TRUE)
if (nrow(essential_genes) == 0) {
  stop("No essential genes found with prop_essential between 0.6 and 0.9")
}

# Section 10: Save essential genes
saveRDS(essential_genes, file = file.path(data_dir, "luad_essential_genes.rds"))

# Section 11: Display results
knitr::kable(head(luad_cells, 10), col.names = "LUAD Cell Line IDs", caption = "Top 10 LUAD Cell Lines")
cat("Total number of LUAD cell lines:", length(luad_cells), "\n")
knitr::kable(head(luad_gene_effect[, 1:6], 10), caption = "Top 10 Rows of LUAD Gene Effect Data (First 5 Genes)")
cat("Total number of LUAD cell lines with gene effect data:", nrow(luad_gene_effect), "\n")
knitr::kable(head(essential_genes, 10), caption = "Top 10 Essential Genes (CERES < -0.5 in 60–90% of LUAD Cell Lines)")
cat("Total number of essential genes:", nrow(essential_genes), "\n")
```

# Candidate Gene Identification

This section identifies candidate target genes by intersecting essential genes from DepMap with significant genes from LINCS and differentially expressed genes from TCGA-LUAD.

```{r, candidate_gene_id, message=FALSE, warning=FALSE}
#| message: false
#| warning: false

# Section 1: Load essential genes
tcga_degs_file <- file.path(data_dir, "tcga_genes_df.rds")
if (!file.exists(tcga_degs_file)) {
  stop("TCGA genes file not found: ", tcga_degs_file)
}
tcga_genes_df <- readRDS(tcga_degs_file)
essential_genes <- readRDS(file.path(data_dir, "luad_essential_genes.rds"))

# Section 2: Define LINCS and TCGA genes
lincs_genes <- sig_gene_symbols
tcga_genes <- tcga_genes_df$Symbol[!is.na(tcga_genes_df$Symbol)]

# Section 3: Find candidate target genes
candidate_targets <- intersect(essential_genes$gene_symbol, intersect(lincs_genes, tcga_genes))
if (length(candidate_targets) == 0) {
  warning("No candidate target genes found in the intersection of DepMap, LINCS, and TCGA")
}

# Section 4: Display results
knitr::kable(data.frame(Candidate_Targets = candidate_targets), caption = "Candidate Target Genes")
cat("Total number of candidate target genes:", length(candidate_targets), "\n")
```

# Visualizations

This section visualizes the TCGA-LUAD differential expression results as a volcano plot, highlighting significant genes.

```{r, visualizations, message=FALSE, warning=FALSE, cache=TRUE}
#| message: false
#| warning: false
#| cache: true

# Section 1: Prepare data for volcano plot
deg_data <- as.data.frame(res) %>%
  dplyr::mutate(
    Gene = tcga_genes[match(rownames(res), tcga_degs_clean)],
    Significant = !is.na(padj) & padj < 0.05 & abs(log2FoldChange) > 1
  )

# Section 2: Create volcano plot
ggplot(deg_data, aes(x = log2FoldChange, y = -log10(padj), color = Significant)) +
  geom_point(size = 1, alpha = 0.5) +
  scale_color_manual(values = c("grey50", "red")) +
  theme_minimal() +
  labs(
    title = "Volcano Plot of TCGA-LUAD Differential Expression",
    x = "Log2 Fold Change (Tumor vs. Normal)",
    y = "-Log10 Adjusted P-value",
    color = "Significant (padj < 0.05, |log2FC| > 1)"
  ) +
  theme(legend.position = "top")}
```

